Ricci-Ollivier Curvature of the Rooted Phylogenetic Subtree-Prune-Regraft Graph

### 2015-07-02 curvature.tex
moved proofs to an appendix

\usepackage{amsmath,amsthm}
\usepackage{csvsimple}
\usepackage{ifthen}
%\usepackage[notref,notcite]{showkeys} % show keys for eqs, etc.
%\usepackage[right]{showlabels}
\usepackage{cite}
\usepackage{enumitem}
\usepackage{url}
\usepackage{subfigure}
\usepackage{graphicx}
\usepackage{todonotes}
\usepackage[nomarkers,nofiglist,notablist]{endfloat}
\let\MYoriglatexcaption\caption
\renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
%\usepackage{todonotes}
%% shift for todonotes
%\usepackage[left=0.5in,right=1.5in,top=1in,bottom=1in,marginparwidth=1.25in]{geometry}
% regular margins
\usepackage[margin=1in]{geometry}

@@

% uncomment to remove the proofs for length estimates
%\usepackage{environ}
%\NewEnviron{killcontents}{}
%\let\proof\killcontents
%\let\endproof\endkillcontents

@@

The automated computational analysis code can be found at \url{https://github.com/matsengrp/curvature}.
\cuttable[We omit most proofs which are unnecessary to understand the paper, and which can be found in~\cite{curvature_arxiv}.]{}

@@

\begin{theorem}
	\label{thm:construct_graph}
	The subgraph of the rSPR graph induced by a set $\mathcal{T}$ of $m$ trees with $n$ leaves can be constructed in $\OhOf{mn^3}$-time.
\end{theorem}
\begin{proof}
	The correctness of the procedure follows by induction on the number of trees already processed, $i$, by observing that the procedure has constructed the subgraph of vertices $1, 2,  \ldots i$ and will construct the subgraph of vertices $1, 2, \ldots i+1$.

@@

\begin{lemma}
	\label{lem:compute_degree}
	For a tree $T$ with $n$ leaves,
	$$\degree{T} = \sum_{u \in T} \size{N(T,u)},$$
	for nodes $u$ of $T$, where $N(T,u)$ is as defined above, and:
	$$\size{N(T,u)} = \begin{cases}
		2n - 1 - x - 4 &\text{if depth($u$) $ > 1$, } \\
		2n - 1 - x - 2 &\text{if depth($u$) $ = 1$ } \\
		0 &\text{if depth($u$) $ \le 0$ } \\
	\end{cases}.$$
	Moreover, each tree of $N(T,u)$ can be obtained by an rSPR operation moving the subtree rooted at $u$.
\end{lemma}
\cuttable{
\begin{proof}
	The statement follows if each of the neighbor assignments are disjoint, that is $N(T,u) \cap N(T,v) = \emptyset$, for all nodes $u$, $v$ of $T$.
	So, suppose, for the purpose of obtaining a contradiction, that there exist two nodes $u$ and $v$ of $T$ such that there exists a tree $S \in (N(T,u) \cap N(T,v))$.
	Then $S$ can be obtained from $T$ by moving the subtrees rooted at $u$ or $v$.
	Call these $U$ and $V$, respectively.
	This implies that both $T \setminus U = S \setminus U$ and $T \setminus V = S \setminus V$ by the definition of an rSPR operation.
	Then the rSPR moves that move $U$ or $V$ to obtain $S$ must be nearest neighbor interchanges (NNIs), that is, rSPR moves which move their subtree to one of four locations: their grandparent edge, aunt edge, sibling's left child edge or sibling's right child edge.
	This implies that, without loss of generality, $U$ is moved to its grandparent edge and $V$ to $U's$ sibling (i.e. (iii) above) or $U$ is moved to its aunt edge and $V$ to $U$'s edge (i.e. (iv) above), a contradiction.
	Therefore the claim holds.
\end{proof}
}

@@

\begin{lemma}
	\label{lem:select_random_neighbor}
	An rSPR neighbor of a tree $T$ can be chosen uniformly at random in $\OhOf{n}$-time\cuttable{ using $\OhOf{n}$ space}.
\end{lemma}
\begin{proof}
	We apply the above procedure.
	We use a standard nodes-and-pointers representation of the trees, which can be constructed in $\OhOf{n}$-time from a Newick string representation and uses linear space in $n$.
	We can compute the degree of $T$ in linear time and space using Lemma~\ref{lem:compute_degree}.
	To efficiently compute $\size{N(T,u)}$ for each node $u$ of $T$, we require the number of nodes $x$ in the subtree rooted at $u$.
	We pre-compute these by (1) labeling each node with its preorder number in a preorder traversal and (2) summing the number of descendant nodes in a postorder traversal and storing the results in an array indexed by preorder number.
	Both of these traversals require $\OhOf{n}$-time.
	There are $2n-1$ = $\OhOf{n}$ nodes of $T$, and $\size{N(T,u)}$ can be computed in constant time using the subtree sizes.
	Moreover, the tree $S$ can be found in $\OhOf{n}$-time by iterating over the edges of $T$ that are not contained within $u$'s subtree to select the corresponding rSPR destination.
	Finally, we require linear time to apply the chosen rSPR operation which entails removing a node, adding a node, and updating a constant number of pointers.
	Thus, the for loop requires linear time.
	By Lemma~\ref{lem:compute_degree} the chosen tree is an rSPR neighbor of $T$ and is chosen uniformly at random.
	Therefore, the procedure uses linear time and space and \cuttable[is correct]{selects an rSPR neighbor of $T$ uniformly at random}.
\end{proof}

@@

We now bound the ratio and difference of rSPR degree between two trees with $n$ leaves.
\begin{lemma}
	\label{lem:degree_max_delta}
	Let $T$ and $S$ be rooted phylogenetic trees with $n \ge 3$ leaves, and assume without loss of generality that $\degree{T} \le \degree{S}$.
	Then:
	\begin{enumerate}
		\item $\frac{\degree{T}}{\degree{S}} \ge 3/4$, and
		\item $\degree{S} - \degree{T} \le n^2 - 5n + 6$.
	\end{enumerate}
\end{lemma}
\cuttable{
\begin{proof}
	To prove (i), we simply note from Lemma~\ref{lem:degree_extremes} that the ladder tree achieves the minimum degree, and the balanced tree achieves the maximum degree:
	\begin{align*}
		\frac{\degree{T}}{\degree{S}} \ge\ &\frac{3n^2 - 13n + 14}{4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)}} \\
		\ge\ &\frac{3n^2 - 13n + 12}{4(n-2)^2 - 2(n-2)} \\
		\cuttable{= \ &\frac{3n^2 - 13n + 12}{4n^2 - 16n + 16 - 2(n-2)} \\}
		= \ &\frac{3n^2 - 13n + 12}{4n^2 - 18n + 20} \\
		\cuttable{\ge\ &\frac{3n^2 - 13n + 12}{4n^2 - 17\frac{1}{3}n + 18} &\forall n \ge 3 \\}
	\end{align*}
	which is greater than 3/4 when $n \ge 3$.
    Similarly for (ii):
	\begin{align*}
		\cuttable{\degree{S} - \degree{T}
		\le\ &(4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)})
		- (3n^2 - 13n + 14) \\}
		\le\ &(4(n-2)^2 - 2(n-2)) - (3n^2 - 13n + 14) \\
		\cuttable{=\ &4n^2 - 16n + 16 - 2n +4 - 3n^2 + 13n - 14 \\}
		=\ &n^2 - 5n + 6.
	\end{align*}
\end{proof}
}

@@

\begin{lemma}
\label{lem:degree_change}
Let $T$ and $S$ be phylogenetic trees such that $S$ can be obtained from $T$ by moving a subtree $R$ with $k$ leaves from its position adjacent to subtree $U$ to a location adjacent to subtree $V$.
Let $L$ be the $\lca{U,V}$ in $T$.
Let $a$ be the number of intermediate nodes on the path from the parent of $R$ to $L$ in $T$, excluding endpoints.
Similarly, let $b$ be the number of intermediate nodes on the path from $V$ to $L$ in $T$, excluding endpoints.
Let $i$ be the number of leaves in $U$ and $j$ be the number of leaves in $V$, excluding any leaves of $R$.
Then the degrees of $T$ and $S$ differ by:
$$2\left(k(a-b) + i - j\right).$$

\end{lemma}
\begin{proof}
The set of permissible rSPR moves changes in four different ways due to the movement of $R$:
(i) subtrees that include nodes on the path from $U$ to $L$ may now be moved into $R$ and its newly introduced parent node,
(ii) subtrees that include nodes on the path from $V$ to $L$ may no longer be moved into $R$ and its parent node,
(iii) $R$'s parent subtree may now be moved into $U$, and
(iv) $R$'s parent subtree may no longer be moved into $V$.
No additional moves are introduced or blocked\cuttable{ by the original rSPR operation on $R$}.

@@

\begin{corollary}
\label{cor:paired_neighbors}
Continuing with the setting and notation in Lemma~\ref{lem:degree_change}, at least
$$o := \deg(T) - 2kb - 2(j-1) = \deg(S) - 2ka - 2(i-1)$$
trees in the neighborhood of $T$ can be paired with $o$ trees in the neighborhood of $S$ such that the pairings are disjoint and $\dspr{T',S'} = 1$ for each $(T',S')$ pair.
\end{corollary}
\cuttable{
\begin{proof}
By the same arguments as in the proof of Lemma~\ref{lem:degree_change}, $o$ rSPR moves can be applied to $T$ and $S$ with the same source and target nodes.
For each such $(T',S')$ pair, we can move $R$ in either tree to obtain the other member of the pair.
\end{proof}
}

We can now use Lemma~\ref{lem:degree_change} to improve the bounds in Lemma~\ref{lem:degree_max_delta} for the case of two adjacent trees.
\begin{lemma}
	\label{lem:degree_max_delta_adjacent}
	Let $T$ and $S$ be rooted phylogenetic trees with $n \ge 3$ leaves, such that $\degree{T} \le \degree{S}$ and $\dspr{T,S} = 1$.
	Then:
	\begin{enumerate}
		\item $\degree{S} - \degree{T} \le 2\floor{\frac{n-2}{2}}\ceil{\frac{n-2}{2}} \le \frac{1}{2} (n-2)^2 $,
		\item $\frac{\degree{T}}{\degree{S}} \ge \frac{5}{6}$, $\forall n \ge 4$, and
		\item $\lim_{n\rightarrow\infty}\frac{\degree{T}}{\degree{S}} =  \frac{6}{7}$.
	\end{enumerate}
\end{lemma}
\cuttable{
\begin{proof}
	We first prove (i).
	By Lemma~\ref{lem:degree_change}, $\degree{S} - \degree{T} = 2(k(a-b) + i - j)$.
	This value is maximized by making $L$ the root and minimizing $b$, namely by setting $b=0$.
	The resulting equation $2(ka + i - j)$ is similarly maximized by setting $i=1$ (which allows us to increase $a$) then maximally balancing the terms in the product $ka$ as follows.

\end{proof}
}

@@

\begin{lemma}
	\label{lem:shared_neighbors}
Let $T$ and $S$ be rooted $X$-trees such that $\dspr{T,S} = 1$.
Then $\size{N(T) \cap N(S)} \le 6n - 17$.
\end{lemma}

\cuttable{
\begin{proof}
	$T$ and $S$ differ by one rSPR move that moves a subtree $R$.
	Pick a neighbor $U \in N(T) \cap N(S)$ of both $T$ and $S$ (this intersection is not empty: $T$ and $S$ are different, so $R$ contains at most $n-2$ of the leaves, thus there must be at least one other tree $U$ obtained by moving $R$ in $T$ and $S$).
	Then either (i) $T$ and $U$ differ in the location of $R$, or (ii) $T$ and $U$ differ in the location of another subtree $Q$.
	In the latter case, $T|(X \setminus L(Q)) = S|(X \setminus L(Q))$ because $T$ and $S$ differ only in the location of $R$ and $\dspr{T,U} = \dspr{S,U} = 1$.
	Then leaves $r' \in R$, $q' \in Q$, and $u' \in U$, for some subtree $U$, form a triple of $T$ and a different triple in $S$.
	This incompatible triple can be resolved in at most $6n - 17$ ways, the maximum of which is reached when $Q$, $U$, and $R$ are themselves a ``triple'' of subtrees.
	By Lemma~\ref{lem:compute_degree}, each of the subtrees is assigned to at most $2n-6$ unique moves.
	Moreover, one additional overlapping move also moves one of the subtrees (that of the aunt of the LCA of the three subtrees).
	The number of shared neighbors is thus at most $3(2n-6) + 1 = 6n-17$.
	Note that this bound is tight when, for example, $T$ and $S$ are ladders with a different configuration of 3 leaves at maximum depth.
\end{proof}
}

@@

\begin{theorem}
Fix a positive integer $k$ and let $R$ be a tree with $k$ leaves.
Let $\{T_n \mid n > k\}$ be a sequence of phylogenetic trees all containing $R$, and let $\{S_n \mid n > k\}$ be the same sequence $T_n$ but with $R$ cut off and attached at a different location.
Then $\lim_{n \rightarrow \infty} \curvature{T_n,S_n} = 0$\cuttable{ for the uniform random walk on the rSPR graph}.
\end{theorem}
\cuttable{
\begin{proof}
Because $d(T_n, S_n) = 1$, we will prove the theorem by showing that the mass transport term $W_{1,n}$ sits between two bounds, each of which has limit 1 as $n$ goes to infinity.

Now we prove the theorem when the number of neighbors differ.
Assume without loss of generality that $\degree{T_n} < \degree{S_n}$.
By Lemma~\ref{lem:degree_change}, $\degree{S_n} - \degree{T_n} = 2(k(a-b) + i - j)$, where each of $\{a,b,i,j\}$ is less than $n$.
Thus, $\degree{S_n} - \degree{T_n} = \OhOf{kn}$.
We again pair neighbor $T'_n$ of $T$ with neighbor $S'_n$ of $S$ such that $\dspr{T'_n, S'_n}=1$ but, as $\degree{T_n} < \degree{S_n}$ we can only account for at most $\degree{T_n} / \degree{S_n}$ of the mass directly and may have to move the $(\degree{S_n} - \degree{T_n}) / \degree{S_n}$ remainder to trees a distance at most 3.
Thus, $W_{1,n}$ is bounded above by $(\degree{T_n} + \OhOf{kn}) / \degree{S_n} = (\degree{S_n} + \OhOf{kn}) / \degree{S_n}$.
We again bound $W_{1,n}$ from below with $(\degree{T_n} - \OhOf{n}) / \degree{T_n}$ by ignoring the mass in common neighbors of $T_n$ and $S_n$.
The theorem again follows because $\degree{T_n}$ is quadratic in $n$.
\end{proof}
}

@@

\begin{lemma}
	\label{lem:curvature_distance_bound}
	Let $T$ and $S$ be two trees. Then:
	$$ \frac{-2}{\dspr{T,S}} \le \curvature{T,S} \le \frac{2}{\dspr{T,S}} .$$
\end{lemma}
\cuttable{
\begin{proof}
	Observe that the distance between neighbors of $T$ and $S$ is bounded between $\dspr{T,S} - 2$ and $\dspr{T,S} + 2$.
	For the curvature upper bound, we then have $\curvature{T,S} \le 1 - \frac{\dspr{T,S} - 2}{\dspr{T,S}} = \frac{2}{\dspr{T,S}}$.
	The lower bound follows similarly.
\end{proof}
}

\begin{lemma}
	The maximum curvature of the uniform random walk between two adjacent trees with $n$ leaves is $\frac{6n-17}{3n^2-13n+14}$.
\end{lemma}
\cuttable{
\begin{proof}
The maximum curvature between adjacent trees $T$ and $S$ occurs when their neighborhoods have maximum overlap and all other tree pairs are at distance 1.
By Lemma~\ref{lem:shared_neighbors} the maximum overlap is $6n-17$.
The amount of overlapping mass in the shared neighbors of $T$ and $S$ is thus $\frac{6n-17}{\max(\degree{T},\degree{S})}$.
The minimum mass transfer cost is thus $1 - \frac{6n-17}{\max(\degree{T},\degree{S})}$.
This is minimized when $\degree{T}=\degree{S}$ are as small as possible, that is $T,S$ are ladders and $\degree{T} = 3n^2 - 13n + 14$.

The maximum curvature is thus $1 - \frac{\degree{T} - (6n-17)}{\degree{T}} = \frac{6n - 17}{\degree{T}}= \frac{6n-17}{3n^2-13n+14}$.
\end{proof}
}
This bound is tight and has been verified computationally for $n \le 7$.

@@

\begin{lemma}
\label{lem:curvature_lower_bound}
	The curvature of the uniform random walk between adjacent trees with $n$ leaves is at least
	$$\frac{-n^2 + 2n}{3.5n^2 - 15n + 16}.$$
\end{lemma}
\cuttable{
\begin{proof}
    In light of Corollary \ref{cor:paired_neighbors}, the optimal mass transport cost is maximized (and therefore curvature minimized) across adjacent trees $T$ and $S$ by a combination of two effects: trees that cannot be paired at distance $1$ and mass that must be moved between unpaired trees due to differing degrees of $T$ and $S$.
		As we will show, these effects can be optimized simultaneously.
    To bound these effects, let $m$ be the maximum (across $T$ and $S$) proportion of mass that cannot be moved between adjacent neighbors of those trees.
    We can bound the mass transport cost from above by $1 + 2m$ because pairs of neighbors of adjacent trees are at most distance 3 apart.
	This gives a lower bound of $1 - (1 + 2m) / 1) = -2m$ on the curvature.

	By Lemmas~\ref{lem:degree_change} and~\ref{lem:degree_max_delta_adjacent}, the latter effect is maximized when the relative degree change is maximized.
	By Corollary~\ref{cor:paired_neighbors}, there are at most $o := \degree{T} -2ka - 2(i-1)$ paired trees, bounding the former effect.
	We now construct a pair of trees that maximizes both effects.
	Let $S$ be the ladder tree with degree $3n^2 - 13n + 14$ and $T$ be the adjacent tree constructed by moving the lower $\floor{\frac{n}{2}}$ leaves of $S$ to the root.
	$T$ has degree at most $3.5n^2 -15n + 16$.
	There are thus $2ka + 2(i-1) = 2\left(\ceil{\frac{n-2}{2}}\floor{\frac{n}{2}} +(1-1)\right) \le \frac{1}{2}n^2 - n$ unpaired neighbors, the maximum possible.
	Moreover, as shown by Lemma~\ref{lem:degree_change} this pair of trees obtains the maximum (absolute and relative) degree change.
	Thus, the maximum $m$ is:
	$$\frac{\frac{1}{2}n^2 - n}{3.5n^2 - 15n + 16}.$$
	The claim follows from multiplying this value by $-2$.
\end{proof}
}

@@

\begin{lemma}
\label{lem:asymptotic}
	Let $T$ and $S$ be rooted phylogenetic trees with $n$ leaves.
	Then:
	\begin{enumerate}
		\item	$\ric{T,S} = \curvature{T,S}$, if $\dspr{T,S} > 1$,
		\item	$\curvature{T,S} \le \ric{T,S} \le \curvature{T,S} + \frac{2}{\max(\degree{T}, \degree{S})}$, if $\dspr{T,S} = 1$.
	\end{enumerate}
\end{lemma}

\cuttable{
\begin{proof}
	We first prove the lower bound in the uniform case, that is $\curvature{T,S} \le \ric{T,S}$.
	Let $W_1(T,S)$ be the mass transport cost in the uniform case, and $W_1'(T,S)$ be the same for the lazy uniform case with parameter $p$.
	Recall that $\curvature{T,S} = \curvature[1]{T,S} = 1 - \frac{W_1(T,S)}{\dspr{T,S}}$, and $\curvature[p]{T,S} / p = \left. \left(1 - \frac{W_1'(T,S)}{\dspr{T,S}}\right) \right/ p$.
	Observe that $$W_1'(T,S) \le pW_1(T,S) + (1-p) \, \dspr{T,S},$$ by the simple mass transport program obtained by treating the mass at $T$ and $S$ as separate from that of the neighbors.
	Then:
	\begin{align*}
		\curvature[p]{T,S} /p &= \left. \left(1 - \frac{W_1'(T,S)}{\dspr{T,S}}\right) \right/ p \\
		&\ge \left. \left(1 - \frac{pW_1(T,S) + (1-p)\dspr{T,S}}{\dspr{T,S}}\right) \right/ p \\
		&= \frac{1}{p} - \frac{W_1(T,S)}{\dspr{T,S}} - \frac{1-p}{p} \\
		&= 1 - \frac{W_1(T,S)}{\dspr{T,S}} \\
		&= \curvature{T,S}.
	\end{align*}

	For the upper bound, we observe that $$W_1'(T,S) \ge p W_1(T,S) + (1-p) \, \dspr{T,S} - \frac{2}{\max(\degree{T}, \degree{S})},$$ as at most $1 / \max(\degree{T},\degree{S})$ of the mass can remain at each of $T$ and $S$, paired with the lazy remainder.
	The upper bound then follows analogously to the lower bound.
	Moreover, no mass can remain at $T$ or $S$ when $\dspr{T,S} > 1$, in which case the curvatures are equal.
\end{proof}
}

@@

\begin{lemma}
	Let $T$ and $S$ be phylogenetic trees with $n$ leaves. Then:
	$$\curvature{T,S} - \frac{1}{3\dspr{T,S}}
	\le \curvature{\MH; T,S}
	\le \curvature{T,S} + \frac{1}{3\dspr{T,S}}, \text{ and }.$$
	$$\curvature{T,S} - 1/6
	\le \curvature{\MH; T,S}
	\le \curvature{T,S} + 1/6$$
\end{lemma}
\cuttable{
\begin{proof}
	We first prove the lower bound.
	By Lemma~\ref{lem:degree_max_delta_adjacent}, the quotient of degrees for two adjacent trees $\ge \frac{5}{6}$.
	Thus, the Hastings ratio is always $\ge \frac{5}{6}$.
	This implies that at most $\frac{1}{6}$ of the mass remains at tree $T$ in the mass distribution.
	Let $W_1(T,S)$ be the cost of an optimal mass transport for the uniform random walk from $T$ to $S$, and $W_1'(T,S)$ the cost for the MH random walk.
	Moreover, let $m_T(z)$ and $m_S(w)$ be the mass assigned for the uniform random walk and $m'_T(z)$ and $m'_S(w)$ be the mass assigned for the MH random walk, for each vertex $z \in N(T)$ and $w \in N(S)$.
	We construct an upper bound on $W_1'(T,S)$ by moving mass according to $W_1$ where possible, and moving the remainder either from $T$ to $S$, from $T$ to a neighbor of $S$, or from a neighbor of $T$ to $S$.
	That is, for each $W_1$ assignment $\xi(z,w)$, we send $\xi'(z,w) = \xi(z,w) \min\left(\frac{m'_T(z)}{m_T(z)}, \frac{m'_S(w)}{m_S(w)}\right)$ of the mass from $z$ to $w$.
	The remaining $\xi(z,w) - \xi'(z,w)$ of the mass is moved from $T$ to $S$, $T$ to $w$, and $z$ to $S$ in the respective proportions
	$\xi(z,w) \max\left(\frac{m'_T(z)}{m_T(z)}, \frac{m'_S(w)}{m_S(w)}\right) - \xi'(z,w)$,
	$\xi(z,w) \min\left(0, \frac{m'_T(z)}{m_T(z)} - \frac{m'_S(w)}{m_S(w)}\right)$, and,
	$\xi(z,w) \min\left(0, \frac{m'_S(w)}{m_S(w)} - \frac{m'_T(w)}{m_T(w)}\right)$.
	The maximum possible mass that is not moved according to $W_1$ is $\frac{1}{6}$.
	Moreover, the affected mass must be moved through at most two additional trees.
	Then, $W_1' \le W_1 + \frac{2}{6}$.
	We now have:
	\begin{align*}
		\curvature{\MH; T,S} & \ge 1 - \frac{W_1 + \frac{1}{3}}{\dspr{T,S}} \\
		& \ge \curvature{T,S} - \frac{1}{3\dspr{T,S}}
	\end{align*}

	We obtain the upper bounds similarly to the lower bounds, by observing that the affected at most $\frac{1}{6}$ of the mass may move through at most two fewer trees (i.e. directly between $T$ and $S$ rather than a pair of neighbors at distance $\dspr{T,S} + 2$ from each other).
	Again, this is at most one fewer tree when $\dspr{T,S}=1$.
\end{proof}
}

@@

\bibliographystyle{IEEEtran}
\bibliography{curvature}
\end{document}

>>>>

\usepackage{amsmath,amsthm}
\usepackage{csvsimple}
\usepackage{ifthen}
%\usepackage[notref,notcite]{showkeys} % show keys for eqs, etc.
%\usepackage[right]{showlabels}
\usepackage{cite}
\usepackage{enumitem}
\usepackage{url}
\usepackage{subfigure}
\usepackage{graphicx}
\usepackage{todonotes}
\usepackage[nomarkers,nofiglist,notablist]{endfloat}
\let\MYoriglatexcaption\caption
\renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
%\usepackage{todonotes}
%% shift for todonotes
%\usepackage[left=0.5in,right=1.5in,top=1in,bottom=1in,marginparwidth=1.25in]{geometry}
% regular margins
\usepackage[margin=1in]{geometry}
\usepackage{thmtools}
\usepackage{thm-restate}

@@

@@

The automated computational analysis code can be found at \url{https://github.com/matsengrp/curvature}.
\cuttable[Proofs of our theorems and lemmas can be found in the appendix.]

@@

%\begin{theorem}
%	\label{thm:construct_graph}
%	The subgraph of the rSPR graph induced by a set $\mathcal{T}$ of $m$ trees with $n$ leaves can be constructed in $\OhOf{mn^3}$-time.
%\end{theorem}
\begin{restatable}{theorem}{constructgraph}
	\label{thm:construct_graph}
	The subgraph of the rSPR graph induced by a set $\mathcal{T}$ of $m$ trees with $n$ leaves can be constructed in $\OhOf{mn^3}$-time.
\end{restatable}

@@

\begin{restatable}{lemma}{computedegree}
	\label{lem:compute_degree}
	For a tree $T$ with $n$ leaves,
	$$\degree{T} = \sum_{u \in T} \size{N(T,u)},$$
	for nodes $u$ of $T$, where $N(T,u)$ is as defined above, and:
	$$\size{N(T,u)} = \begin{cases}
		2n - 1 - x - 4 &\text{if depth($u$) $ > 1$, } \\
		2n - 1 - x - 2 &\text{if depth($u$) $ = 1$ } \\
		0 &\text{if depth($u$) $ \le 0$ } \\
	\end{cases}.$$
	Moreover, each tree of $N(T,u)$ can be obtained by an rSPR operation moving the subtree rooted at $u$.
\end{restatable}

@@

\begin{restatable}{lemma}{selectrandomneighbor}
	\label{lem:select_random_neighbor}
	An rSPR neighbor of a tree $T$ can be chosen uniformly at random in $\OhOf{n}$-time\cuttable{ using $\OhOf{n}$ space}.
\end{restatable}

@@

We now bound the ratio and difference of rSPR degree between two trees with $n$ leaves.
\begin{restatable}{lemma}{degreemaxdelta}
	\label{lem:degree_max_delta}
	Let $T$ and $S$ be rooted phylogenetic trees with $n \ge 3$ leaves, and assume without loss of generality that $\degree{T} \le \degree{S}$.
	Then:
	\begin{enumerate}
		\item $\frac{\degree{T}}{\degree{S}} \ge 3/4$, and
		\item $\degree{S} - \degree{T} \le n^2 - 5n + 6$.
	\end{enumerate}
\end{restatable}

@@

\begin{restatable}{lemma}{degreechange}
\label{lem:degree_change}
Let $T$ and $S$ be phylogenetic trees such that $S$ can be obtained from $T$ by moving a subtree $R$ with $k$ leaves from its position adjacent to subtree $U$ to a location adjacent to subtree $V$.
Let $L$ be the $\lca{U,V}$ in $T$.
Let $a$ be the number of intermediate nodes on the path from the parent of $R$ to $L$ in $T$, excluding endpoints.
Similarly, let $b$ be the number of intermediate nodes on the path from $V$ to $L$ in $T$, excluding endpoints.
Let $i$ be the number of leaves in $U$ and $j$ be the number of leaves in $V$, excluding any leaves of $R$.
Then the degrees of $T$ and $S$ differ by:
$$2\left(k(a-b) + i - j\right).$$

\end{restatable}

@@

\begin{restatable}{corollary}{pairedneighbors}
\label{cor:paired_neighbors}
Continuing with the setting and notation in Lemma~\ref{lem:degree_change}, at least
$$o := \deg(T) - 2kb - 2(j-1) = \deg(S) - 2ka - 2(i-1)$$
trees in the neighborhood of $T$ can be paired with $o$ trees in the neighborhood of $S$ such that the pairings are disjoint and $\dspr{T',S'} = 1$ for each $(T',S')$ pair.
\end{restatable}

We can now use Lemma~\ref{lem:degree_change} to improve the bounds in Lemma~\ref{lem:degree_max_delta} for the case of two adjacent trees.
\begin{restatable}{lemma}{degreemaxdeltaadjacent}
	\label{lem:degree_max_delta_adjacent}
	Let $T$ and $S$ be rooted phylogenetic trees with $n \ge 3$ leaves, such that $\degree{T} \le \degree{S}$ and $\dspr{T,S} = 1$.
	Then:
	\begin{enumerate}
		\item $\degree{S} - \degree{T} \le 2\floor{\frac{n-2}{2}}\ceil{\frac{n-2}{2}} \le \frac{1}{2} (n-2)^2 $,
		\item $\frac{\degree{T}}{\degree{S}} \ge \frac{5}{6}$, $\forall n \ge 4$, and
		\item $\lim_{n\rightarrow\infty}\frac{\degree{T}}{\degree{S}} =  \frac{6}{7}$.
	\end{enumerate}
\end{restatable}

@@

\begin{restatable}{lemma}{sharedneighbors}
	\label{lem:shared_neighbors}
Let $T$ and $S$ be rooted $X$-trees such that $\dspr{T,S} = 1$.
Then $\size{N(T) \cap N(S)} \le 6n - 17$.
\end{restatable}

@@

\begin{restatable}{theorem}{flatcurvature}
\label{thm:flat_curvature}
Fix a positive integer $k$ and let $R$ be a tree with $k$ leaves.
Let $\{T_n \mid n > k\}$ be a sequence of phylogenetic trees all containing $R$, and let $\{S_n \mid n > k\}$ be the same sequence $T_n$ but with $R$ cut off and attached at a different location.
Then $\lim_{n \rightarrow \infty} \curvature{T_n,S_n} = 0$\cuttable{ for the uniform random walk on the rSPR graph}.
\end{restatable}

@@

\begin{restatable}{lemma}{curvaturedistancebound}
	\label{lem:curvature_distance_bound}
	Let $T$ and $S$ be two trees. Then:
	$$ \frac{-2}{\dspr{T,S}} \le \curvature{T,S} \le \frac{2}{\dspr{T,S}} .$$
\end{restatable}

\begin{restatable}{lemma}{maxadjacentcurvature}
	\label{lem:max_adjacent_curvature}
	The maximum curvature of the uniform random walk between two adjacent trees with $n$ leaves is $\frac{6n-17}{3n^2-13n+14}$.
\end{restatable}
This bound is tight and has been verified computationally for $n \le 7$.

@@

\begin{restatable}{lemma}{curvaturelowerbound}
\label{lem:curvature_lower_bound}
	The curvature of the uniform random walk between adjacent trees with $n$ leaves is at least
	$$\frac{-n^2 + 2n}{3.5n^2 - 15n + 16}.$$
\end{restatable}

@@

\begin{restatable}{lemma}{asymptotic}
\label{lem:asymptotic}
	Let $T$ and $S$ be rooted phylogenetic trees with $n$ leaves.
	Then:
	\begin{enumerate}
		\item	$\ric{T,S} = \curvature{T,S}$, if $\dspr{T,S} > 1$,
		\item	$\curvature{T,S} \le \ric{T,S} \le \curvature{T,S} + \frac{2}{\max(\degree{T}, \degree{S})}$, if $\dspr{T,S} = 1$.
	\end{enumerate}
\end{restatable}

@@

\begin{restatable}{lemma}{mhcurvature}
	\label{lem:mh_curvature}
	Let $T$ and $S$ be phylogenetic trees with $n$ leaves. Then:
	$$\curvature{T,S} - \frac{1}{3\dspr{T,S}}
	\le \curvature{\MH; T,S}
	\le \curvature{T,S} + \frac{1}{3\dspr{T,S}}, \text{ and }.$$
	$$\curvature{T,S} - 1/6
	\le \curvature{\MH; T,S}
	\le \curvature{T,S} + 1/6$$
\end{restatable}

@@

\bibliographystyle{IEEEtran}
\bibliography{curvature}

\appendices

\section{Supplementary Proofs}

\constructgraph*
\begin{proof}
	The correctness of the procedure follows by induction on the number of trees already processed, $i$, by observing that the procedure has constructed the subgraph of vertices $1, 2,  \ldots i$ and will construct the subgraph of vertices $1, 2, \ldots i+1$.

\computedegree*
\begin{proof}
	The statement follows if each of the neighbor assignments are disjoint, that is $N(T,u) \cap N(T,v) = \emptyset$, for all nodes $u$, $v$ of $T$.
	So, suppose, for the purpose of obtaining a contradiction, that there exist two nodes $u$ and $v$ of $T$ such that there exists a tree $S \in (N(T,u) \cap N(T,v))$.
	Then $S$ can be obtained from $T$ by moving the subtrees rooted at $u$ or $v$.
	Call these $U$ and $V$, respectively.
	This implies that both $T \setminus U = S \setminus U$ and $T \setminus V = S \setminus V$ by the definition of an rSPR operation.
	Then the rSPR moves that move $U$ or $V$ to obtain $S$ must be nearest neighbor interchanges (NNIs), that is, rSPR moves which move their subtree to one of four locations: their grandparent edge, aunt edge, sibling's left child edge or sibling's right child edge.
	This implies that, without loss of generality, $U$ is moved to its grandparent edge and $V$ to $U's$ sibling (i.e. (iii) above) or $U$ is moved to its aunt edge and $V$ to $U$'s edge (i.e. (iv) above), a contradiction.
	Therefore the claim holds.
\end{proof}

\selectrandomneighbor*
\begin{proof}
	We apply the above procedure.
	We use a standard nodes-and-pointers representation of the trees, which can be constructed in $\OhOf{n}$-time from a Newick string representation and uses linear space in $n$.
	We can compute the degree of $T$ in linear time and space using Lemma~\ref{lem:compute_degree}.
	To efficiently compute $\size{N(T,u)}$ for each node $u$ of $T$, we require the number of nodes $x$ in the subtree rooted at $u$.
	We pre-compute these by (1) labeling each node with its preorder number in a preorder traversal and (2) summing the number of descendant nodes in a postorder traversal and storing the results in an array indexed by preorder number.
	Both of these traversals require $\OhOf{n}$-time.
	There are $2n-1$ = $\OhOf{n}$ nodes of $T$, and $\size{N(T,u)}$ can be computed in constant time using the subtree sizes.
	Moreover, the tree $S$ can be found in $\OhOf{n}$-time by iterating over the edges of $T$ that are not contained within $u$'s subtree to select the corresponding rSPR destination.
	Finally, we require linear time to apply the chosen rSPR operation which entails removing a node, adding a node, and updating a constant number of pointers.
	Thus, the for loop requires linear time.
	By Lemma~\ref{lem:compute_degree} the chosen tree is an rSPR neighbor of $T$ and is chosen uniformly at random.
	Therefore, the procedure uses linear time and space and \cuttable[is correct]{selects an rSPR neighbor of $T$ uniformly at random}.
\end{proof}
